Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-12-20 paint-contour-tree.tex
update for dcg

\InNotSoCGVer{
\author{
  Benjamin Raichel%\BenThanks{}
  \and
  C. Seshadhri%\thanks{University of California, Santa Cruz; {\tt scomandu@ucsc.edu}}
}

@@

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
If $C$ denotes the set of critical points in $\MM$, the running time is
roughly $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
or in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
An edge of the contour tree connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.) 
\Fig{relative} and \Fig{sorting} 
show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards. Leaves are either maxima or minima,
and internal nodes are either ``joins" or ``splits".

@@

Essentially, the ``run time per critical point" is determined by the height/depth of the point in the contour tree.
This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
where $D$ is the diameter of the contour tree. 
This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
then we get a bound of $O(t\log\log t)$.
Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.

@@

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value $t=\Omega(N)$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $t\log t$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $\thmConst$ comparisons. Then \emph{any algorithm} must make roughly $\thmConst$
comparisons in the worst-case over this set. (Further details are in \InSoCGVer{the full version \CiteFullVer}\InNotSoCGVer{\Sec{lb}}.)

@@

We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. So $f$ is explicitly defined only on the vertices of $\MM$,
and all other values are obtained by linear interpolation.

@@

\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x| f(x) = h\}$.
A \emph{contour} is a connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
\end{definition}

@@

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.
}
\InNotSoCGVer{
\subsection{Some technical remarks}
\technicalRemarks
}

@@

\newcommand{\intuitionSection}{
\section{A tour of the new contour tree algorithm} \label{sec:approach}
\InNotSoCGVer{
Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
}
\InSoCGVer{
For the $d=2$ case, where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
can be explained clearly.  Therefore, here we provide a high level view of the entire result.
}

@@

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
(Our real definition is more complicated, and involves simplicial complexes
that allow additional ``trivial'' minima.)
In such a complex, there exists a non-ascending path from any point to this unique minimum.
Analogously, we can define maximum dominant complexes,
and both are collectively called extremum dominant.

We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
One way to think of our procedure is a meteorological analogy. Take an arbitrary maximum $x$,
and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
while there may be numerous disconnected dry parts. The interface between the dry and wet parts
is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
given by the ``water line". The wet part is clearly maximum dominant, since all
wet points have a non-descending path to $x$. So we can simply cut along the interface
contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
we are left with a set of disconnected dry parts, and we can recur this procedure on them.

@@

We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-03},
one may think that it suffices to design faster algorithms for join trees. But this
cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
to construct the true join tree must construct the path of splits, which implies
sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
of building the join tree can be related to that of building $\cC(\MM)$.

@@

The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
paint the interior of higher dimensional faces.} That is, this process is restricted to the 1-skeleton of $\NN$.
Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
paint to reach it.  In the following, $[n]$ denotes the set $\{1, \ldots, n\}$, for any natural number $n$.

@@

This is precisely how we hope to exploit the information in the painting. We prove later that
when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
data structure for that.

@@

In this discussion, we ignored an annoying problem. Vertices
may actually be touched by numerous colors, not just one or two as assumed above. A simple solution
would be to insert vertices into heaps corresponding to all colors touching it. But there could
be super-constant numbers of copies of a vertex, and handling all these copies would lead
to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps, one for each
``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges have
at most $2$ colors, because of all the color merges that previously occurred.

@@

The intuition is that the paint spilling from the maxmima in the simplicial complex, corresponds to paint spilling from the leaves in 
the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
higher than the far left one, it will merge into the component containing the blue mound (and also the yellow and red mounds)
before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the join tree.

@@

\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
is non-empty and consists of $\phi_1, \ldots, \phi_k$ (as denoted in the procedure).
By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
Consider $\wet(x,\NN_{k+1})$. The interface is exactly $\phi_1, \ldots, \phi_k$.
So the only dry vertices are those in the boundaries $B_1, \ldots, B_k$. But these
boundaries are maxima.
\end{proof}

@@

\begin{lemma}\label{lem:new-verts}
A face $F \in \MM$ is cut\footnote{Technically what we are calling a single cut is done with two hyperplanes.} at most once during $\rain(x,\MM)$.
\end{lemma}
\begin{proof} Notation here follows the pseudocode of $\rain$.
First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
Second, all recursive calls are made on dry complexes.

@@

Using the above lemma we can now prove that computing the contour tree for a minimum dominant complex 
amounts to computing its join tree.  Specifically, to prove our main theorem, we rely on the correctness of the
merging procedure from~\cite{csa-cctad-03} that constructs the contour tree from the join and split trees. 
It actually constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
obtained by replacing each edge in the contour tree with a path of all regular vertices 
(sorted by height) whose corresponding contour belongs to the equivalence class of that edge.

Consider a tree $T$ with a vertex $v$ of in and out degree at most $1$.
\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, 
delete $v$ and connect its neighbors by an edge (i.e.\ smooth $v$ out). This tree is denoted by $T \ominus v$.

@@

\smallskip
\begin{compactenum}
	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
	\item Denote $v$ as a \emph{candidate} if the sum of its indegree in $\cJ$ and outdegree in $\cS$ is $1$. 
	\item Add all candidates to queue.
	\item While candidate queue is non-empty: 
	\begin{compactenum}
		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
		consider its edge in $\cS$. In either case, denote the edge by $(v,w)$.
		\item Insert $(v,w)$ in $\cA(\MM)$. 
		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
	\end{compactenum}
	\item Smooth out all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
\end{compactenum}
\end{minipage}}

@@

\begin{definition} 
\label{def:criticalJoin}
The \emph{critical join tree} $\jc(\MM)$ is built on the set $V'$ of all
critical points other than the non-dominant minima. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
and $v$ is adjacent (in $\MM$) to a vertex in this component. \InSoCGVer{The \emph{join tree}, $\cJ(\MM)$, is defined analogously, but instead on $\cV(\MM)$.}
%
%\InSoCGVer{Have sentence defining the join tree here.}
%
\end{definition}

@@

\newcommand{\proofofContourJoinEquiv}{
\begin{proof} We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
maxima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
The join tree is now $\cJ(\MM) \ominus \{m_i\}$. We can now process $\cJ$ leaf by leaf, and all
edges of $\cJ$ are inserted into $\cA(\MM)$.

Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Similarly, smoothing out regular
points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
\end{proof}
}

@@

\begin{Remark}\label{rem:joinAndCriticalJoin}
The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
it suffices to have an efficient algorithm for computing $\jc(\MM)$.
Due to minor technicalities, it is easier to phrase the following section instead in terms of computing $\cJ(\MM)$ efficiently.  
Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
\end{Remark}

@@

\medskip
\noindent
{\bf Color list $\col(v)$:} For each point $v$, we maintain $\col(v)$ as a simple list.
In addition, we will maintain another (sub)list of colors $L$ such that $\forall c \in L$,
$v$ is \emph{not} the highest vertex in $T(\rep(c))$. Note that given $c \in \col(v)$
and $\rep(c)$, this property can be checked in constant time. If $c$ is ever removed
from this sublist, it can never enter it again.

@@

We formally describe the algorithm below. 
We require a technical definition of \emph{ripe} vertices.

@@

\smallskip
\begin{compactenum}
	\item Run $\init(\MM)$.
	\item While there are unprocessed critical points:
	\begin{compactenum}
		\item Run $\update(\stack)$. Pop $\stack$ to get $h$.
		\item Let $\cur(h) = \{\rep(c) | c \in \col(h)\}$.
		\item For all $c' \in \cur(h)$:
		\begin{compactenum}
			\item Add edge $(\h(c'),h)$ to $\cJ(\MM)$.
			\item Delete $h$ from $T(c')$.
		\end{compactenum}
	\item Merge heaps $\{T(c') | c' \in \cur(h)\}$.
	\item Take union of $\cur(h)$ and denote resulting color by $\widehat{c}$.
	\item Set $\h(\widehat{c}) = h$ and mark $h$ as processed.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

A few simple facts:
\begin{compactitem}
	\item At all times, the colors form a valid painting.
	\item Each vertex is present in at most $2$ heaps. After processing, it is removed from all heaps.
	\item After $v$ is processed, all edges incident to $v$ have the same (representative) color. %(more technically, the same representative). 
%	\item When $\update(\stack)$ terminates, the head $h$ is ripe. 
	\item Vertices on the stack are in increasing height order. 
\end{compactitem}

@@

Our main workhorse is the following technical lemma. In the following, the current color of an edge, $e$, is the value of $\rep(\chi(e))$, 
where $\chi(e)$ is the color of $e$ from the initial painting.

@@

\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
by an edge $e$ which is currently colored $c$. At all times: either all edges in $\PP$ are currently colored $c$, or there
exists a critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
\end{lemma}

\begin{proof} Since $e$ has color $c$,
there must exist vertices in $\PP$ touched by $c$. Consider the highest
vertex $w$ in $\PP$ that is touched by $c$ and some other color. If no such vertex exists,
this means all edges incident to a vertex touched by $c$ are colored $c$. By walking through
$\PP$, we deduce that all edges are colored $c$.

>>>>

\InNotSoCGVer{
\author{
  Benjamin Raichel\thanks{%
      Department of Computer Science; %
      University of Texas at Dallas; %
      Richardson, TX, 75080, USA; %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
  }
  \and
  C. Seshadhri\thanks{%
        Department of Computer Science; %
      University of California, Santa Cruz; %
      Santa Cruz, CA, 95064, USA; %
      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
  }
}

@@

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
If $C$ denotes the set of critical points in $\MM$, the running time is
roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
or in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
An edge of the contour tree connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. We provide formal definitions later. 
\Fig{relative} and \Fig{sorting} 
show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards. Leaves are either maxima or minima,
and internal nodes are either ``joins" or ``splits".

@@

Essentially, the ``run time per critical point" is determined by the height/depth of the point in the contour tree.
This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
where $D$ is the directed diameter of the contour tree. 
This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
then we get a bound of $O(t\log\log t)$.
Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.

@@

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value $t=\Omega(N)$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $c t\log t$, for some constant $c$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $\alpha \thmConst$ comparisons, for any constant $\alpha$. Then \emph{any algorithm} must make roughly $\thmConst$
comparisons in the worst-case over this set. (Further details are in \InSoCGVer{the full version \CiteFullVer}\InNotSoCGVer{\Sec{lb}}.)

@@

We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. We assume that all vertices of $\MM$ have distinct function values, 
and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.

@@

\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x\in \MM| f(x) = h\}$.
A \emph{contour} is a connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
\end{definition}

@@

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every regular vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.
}
\InNotSoCGVer{
\subsection{Some remarks}
\technicalRemarks
}

@@

\newcommand{\intuitionSection}{
\section{A tour of the new contour tree algorithm} \label{sec:approach}
Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
can be explained clearly.  Therefore, here we first provide a high level view of the entire result.

@@

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
(Our real definition is more complicated, and involves simplicial complexes
that allow additional ``trivial'' minima.)
In such a complex, there exists a non-ascending path from any point to this unique minimum.
Analogously, we can define \emph{maximum dominant} complexes,
and both are collectively called \emph{extremum dominant}.

We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
One way to think of our procedure is a meteorological analogy. Take an arbitrary maximum $x$,
and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
while there may be numerous disconnected dry parts. The interface between the dry and wet parts
%is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
is a set of limits of sequences of contours, which informally we call just a set of contours, which are
given by the ``water line". The wet part is clearly maximum dominant, since all
wet points have a non-descending path to $x$. So we can simply cut along the interface
contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
we are left with a set of disconnected dry parts, and we can recur this procedure on them.

@@

We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-03},
one may think that it suffices to design faster algorithms for join trees. But this
cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
to construct the true join tree must construct the path of splits, which implies
sorting all of them. Thus it is necessary to cut $\MM$ into pieces where the cost
of building the join tree can be related to that of building $\cC(\MM)$.

@@

The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
paint the interior of faces with dimension strictly larger than 1.} That is, this process is restricted to the 1-skeleton of $\NN$.
Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
paint to reach it.  In the following, $[n]$ denotes the set $\{1, \ldots, n\}$, for any natural number $n$.

@@

This is precisely how we hope to exploit the information in the painting. We prove later that
when some join $v$ has exactly two colors among all incident edges with $v$ as the lower endpoint, the corresponding maxima (of those colors) are exactly
the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
data structure for that. So the red and blue mounds are not actually recolored purple, 
and instead say red now points to blue in the union find data structure.

@@

In this discussion, we ignored an annoying problem. Vertices
may actually be touched by numerous colors, not just one or two as assumed above. A simple solution
would be to insert vertices into heaps corresponding to all colors touching it. But there could
be super-constant numbers of copies of a vertex, and handling all these copies would lead
to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps, one for each
``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges incident from above have
at most $2$ colors, because of all the color merges that previously occurred.

@@

The intuition is that the paint spilling from the maxmima in the simplicial complex, corresponds to paint spilling from the leaves in 
the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
higher than the far left one, the merge event involving the component containing the blue mound (and also the yellow and red mounds) and corresponding to the far right saddle
will occur before the one corresponding to the far left saddle.  Hence, the vertices initially touched by blue are not contiguous in the join tree.

@@

\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
is non-empty and consists of $\phi_1, \ldots, \phi_k$ (as denoted in the procedure).
By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
Consider $\wet(x,\NN_{k+1})$. The interface is exactly $\phi_1, \ldots, \phi_k$.
So the only dry vertices of $\NN_{k+1}$ are those on the boundaries created by calling $\cut$ for each $\phi_i$, 
but such boundaries are all maxima.
%$B_1, \ldots, B_k$. But these boundaries are maxima.
\end{proof}

@@

\begin{lemma}\label{lem:new-verts}
A face $F \in \MM$ is cut\footnote{What we are calling a single cut is actually done with two hyperplanes.} at most once during $\rain(x,\MM)$.
\end{lemma}
\begin{proof} Notation here follows the pseudocode of $\rain$.
First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
Second, all recursive calls are made on dry complexes.

@@

Using the above lemma we can now prove that computing the contour tree for a minimum dominant complex 
amounts to computing its join tree.  Specifically, to prove our main theorem, we rely on the correctness of the
merging procedure from~\cite{csa-cctad-03} that constructs the contour tree from the join and split trees. 
It actually first constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
obtained by replacing each edge in the contour tree with a path of all regular vertices 
(sorted by height) whose corresponding contour belongs to the equivalence class of that edge.

Consider a tree $T$ with a vertex $v$ of in and out degree at most $1$.
\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, 
delete $v$ and connect its neighbors by an edge (i.e.\ smooth $v$ out). This tree is denoted by $T \ominus v$.
%
The merging procedure of \cite{csa-cctad-03} is shown below as $\merge(\cJ(\MM),\cS(\MM))$.  
This procedure builds $\cA(\MM)$ by iteratively adding edges that are adjacent to leaves of either $\cJ(\MM)$ or $\cS(\MM)$, 
erasing the corresponding vertex from both $\cJ(\MM)$ and $\cS(\MM)$, and repeating.  $\cC(\MM)$ is then produced by erasing all regular vertices in $\cA(\MM)$.
We use the fact that this procedure correctly produces the contour tree (regardless of the order leaves are processed), 
and refer the interested reader to \cite{csa-cctad-03} for the proof of this fact, as it is not needed for our purposes.

@@

\smallskip
\begin{compactenum}
	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
	\item Denote $v$ as a \emph{candidate} if the sum of its indegree in $\cJ$ and outdegree in $\cS$ is $1$. 
	\item Add all candidates to queue.
	\item While candidate queue is non-empty: 
	\begin{compactenum}
		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
		consider its edge in $\cS$. In either case, denote the edge by $(v,w)$.
		\item Insert $(v,w)$ in $\cA(\MM)$. 
		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
	\end{compactenum}
	\item Erase all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
\end{compactenum}
\end{minipage}}

@@

\begin{definition} 
\label{def:criticalJoin}
The \emph{critical join tree} $\jc(\MM)$ is built on the set $V'$ of all
critical points other than the non-dominant minima. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
and $v$ is adjacent (in $\MM$) to a vertex in this component. \InSoCGVer{The \emph{join tree}, $\cJ(\MM)$, is defined analogously, but instead on $\cV(\MM)$.}
\end{definition}

@@

\newcommand{\proofofContourJoinEquiv}{
\begin{proof} %We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
minima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
Moreover the join tree is now $\cJ(\MM) \ominus \{m_i\}$, and so we can now process $\cJ$ leaf by leaf, and all
remaining edges of $\cJ$ are inserted into $\cA(\MM)$.
%
Note that smoothing out regular points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
Thus after smoothing out all regular points from $\cA(\MM)$ we recover the edges of $\jc(\MM)$.  
Hence the resulting contour tree contains these edges and the $\{(s_i,m_i)\}$ as claimed.
%Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Similarly, smoothing out regular
%points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
\end{proof}
}

@@

\begin{Remark}\label{rem:joinAndCriticalJoin}
The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
it suffices to have an efficient algorithm for computing $\jc(\MM)$.
In the other sections of the paper it is better to phrase things instead in terms of computing $\cJ(\MM)$ efficiently.  
%Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
%with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
Note however that for minimum dominant complexes, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.  
(Indeed $\jc(\MM)$ was only defined to make the above theorem statement clearer.)
\end{Remark}

@@

\medskip
\noindent
{\bf Color list $\col(v)$:} For each vertex $v$, we maintain $\col(v)$ as a simple list.
In addition, we will maintain another (sub)list of colors $L$ such that $\forall c \in L$,
$v$ is \emph{not} the highest vertex in $T(\rep(c))$. Note that given $c \in \col(v)$
and $\rep(c)$, this property can be checked in constant time. If $c$ is ever removed
from this sublist, it can never enter it again.

@@

We formally describe the algorithm below. 
We require the following definition of \emph{ripe} vertices.

@@

\smallskip
\begin{compactenum}
	\item Run $\init(\MM)$.
	\item While there are unprocessed critical points:
	\begin{compactenum}
		\item Run $\update(\stack)$. Pop $\stack$ to get $h$.
		\item Let $\cur(h) = \{\rep(c) | c \in \col(h)\}$.
		\item For all $c' \in \cur(h)$:
		\begin{compactenum}
			\item Add edge $(\h(c'),h)$ to $\cJ(\MM)$.
			\item Delete $h$ from $T(c')$.
		\end{compactenum}
	\item Merge heaps $\{T(c') | c' \in \cur(h)\}$.
	\item Take union of $\cur(h)$ and denote resulting representative color by $\widehat{c}$.
	\item Set $\h(\widehat{c}) = h$ and mark $h$ as processed.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

A few simple facts:
\begin{compactitem}
	\item At all times, the colors form a valid painting.
	\item Each vertex is present in at most $2$ heaps. After processing, it is removed from all heaps.
	\item After $v$ is processed, all edges incident to $v$ from above have the same (representative) color. %(more technically, the same representative). 
%	\item When $\update(\stack)$ terminates, the head $h$ is ripe. 
	\item Vertices on the stack are in increasing height order. 
\end{compactitem}

@@

Our main workhorse is the following lemma. The \emph{current color} of an edge, $e$, refers to the value of $\rep(\chi(e))$, 
where $\chi(e)$ is the color of $e$ from the initial painting.

@@

\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
by an edge $e$ which is currently colored $c$. At all times: either all edges in $\PP$ are currently colored $c$, or there
exists a critical vertex $w \in \PP$ currently fully touched by $c$ and currently touched by another color.
\end{lemma}

\begin{proof} Since $e$ has current color $c$,
there must exist vertices in $\PP$ currently touched by $c$. Consider the highest
vertex $w$ in $\PP$ that is currently touched by $c$ and some other color. If no such vertex exists,
this means all edges incident to a vertex currently touched by $c$ are currently colored $c$. By walking through
$\PP$, we deduce that all edges are currently colored $c$.
